# Basic information

## What is a model?


## What is model validation?
In most basic terms, model validation is to answer the question, "is this model good?" That can have different meanings in different contexts.

Validation focuses on ensuring that the model description is accurate to its performance. Another way to think of it is checking model assumptions. 


# Essential concepts

## Fitting vs. prediction


## In-sample vs out-of-sample


## Fitted vs. predicted values

```{r residual-types, echo=FALSE }
data(cars)

# indicate the training points
indx = cars$speed >=15 & cars$speed <=17

# fit a spline with too much wiggliness to the points
xx = seq( min(cars$speed), max(cars$speed), length.out=200 )
spl = smooth.spline(cars$speed[!indx], cars$dist[!indx], spar=0.02)

# fit a second-order polynomial to the points
lm1 = lm( dist ~ 0 + poly(speed, 2), data=cars )

# plot the training points and the smoothing spline
plot( cars[!indx, ], bty='n' )
```


```{r residual-types, echo=FALSE }
# plot the training points and the smoothing spline
plot( cars[!indx, ], bty='n' )
lines( predict(spl, xx) )
lines( x=xx, y=mean(cars$dist[!indx]) + predict(lm1, data.frame(speed = xx)) )

# add the held-out points to the plot, highlighting the predictive residuals
points( cars[indx, ], col='red')
```

# Model types

We are going to use a couple of example models throughout this workshop. First, a logistic regression model and second, a random forest regression model.



## Logistic regression
Logistic regression is a method of estimating 




# Import data

Let's import the data for our examples. 


