## In-sample vs. out-of-sample
In-sample and out-of-sample describe data relative to some model. The data are in-sample to the model if the data were used in training that model, and are out-of-sample otherwise.


## Fitting vs. prediction
That leads to the key difference between fitting and prediction: fitting is done with in-sample data, and prediction is done with out-of-sample data. 

```{r residual-types, echo=FALSE }
data( cars )

# indicate the training points
indx = cars$speed >=15 & cars$speed <=17

# fit a spline with too much wiggliness to the points
xx = seq( min(cars$speed), max(cars$speed), length.out=200 )
spl = smooth.spline(cars$speed[!indx], cars$dist[!indx], spar=0.02)

# fit a second-order polynomial to the points
lm1 = lm( dist ~ 0 + poly(speed, 2), data=cars )

# plot the training points and the smoothing spline
plot( cars[!indx, ], bty='n' )
lines( xx, mean(cars$dist) + predict(lm1, list(speed=xx)), lty=2 )
```


```{r echo=FALSE }
# plot the training points and the smoothing spline
plot( cars[!indx, ], bty='n' )
lines( predict(spl, xx) )
lines( x=xx, y=mean(cars$dist[!indx]) + predict(lm1, data.frame(speed = xx)) )

# add the held-out points to the plot, highlighting the predictive residuals
points( cars[indx, ], col='red' )
```
